---
title: Avoiding integrals in statistics
author: David J. Harris
date: '2017-08-22'
categories:
  - Approximation
  - Integration
  - Latent variables
  - Monte Carlo
  - Statistics
  - Variational methods
slug: avoiding-integrals-in-statistics
---


```{r setup, include=FALSE}
library(tidyverse)
library(cowplot)
library(viridis)
library(rethinking)
knitr::opts_chunk$set(echo = TRUE)
```

# Why integrate?

independent random events whose intensity is drawn from a gamma distribution.

, so we can just use
`dnbinom` in R, or `scipy.stats.nbinom.pmf` in python. But what if 

```{r}
set.seed(1)
N = 1000
shape = 4
rate = 1
xmax = 12
posterior_height = .04
```

```{r}
curve(dgamma(x, shape, rate), to = xmax, yaxs = "i", bty = "l",
      n = 1000, ylim = c(0, 1), ylab = "Probability")
prob = 1 / (1 + 1/rate)
lines(0:100, dnbinom(0:100, size = shape, prob = prob), col = "darkred",
      type = "h")
lines(0, dnbinom(0, size = shape, prob = prob), type = "h", col = 2, lwd = 3)
```

```{r}
curve(dgamma(x, shape, rate), to = xmax, yaxs = "i", bty = "l",
      n = 1000, ylim = c(0, 1), ylab = "Probability")
curve(dpois(0, x), to = xmax, add = TRUE, col = 2)
arrows(5, .4, 2, .2, lwd = 2, length = .125)
```
If the intensity has a typical value lik 5 or 10, we *could* still get unlucky 
and get nothing, but it wouldn't be very likely. Alternatively, if the intensity
is less than 1, then we wouldn't expect anything, but those extremely low 
intensities are pretty rare. And what about all the values in between?

```{r}
likelihood = function(lambda){
  dpois(0, lambda) * dgamma(lambda, shape, rate)
}
curve(likelihood, to = xmax, yaxs = "i", bty = "l",
      n = 1000, ylim = c(0, posterior_height), ylab = "Probability")
```

This gives us all the ways that we could have gotten zero, weighted by their 
probability. If we want the total probability associated with zero, we "just" 
need to add up those weights. Adding up all the possible weights gives us
this integral

$$\int_0^\infty p(k=0|\lambda)~~p(\lambda|\alpha\beta)~~d\lambda,$$

where
$p(k|\lambda)=\frac{\lambda^k e^{-\lambda}}{k!}$ (i.e. the Poisson likelihood) and $p(\lambda|\alpha,\beta)=\frac{\beta^\alpha}{\Gamma(\alpha)}x^{\alpha-1}e^{-\beta \lambda}$ (i.e. the gamma likelihood).

Fortunately, I chose an example where this integral is known: it's "just" the 
negative binomial distribution's probability mass function, i.e. 

$$p(k|p,\alpha)={{k + \alpha - 1}\choose{k}} (1-p)^\alpha p^k.$$

where $p=1 / (1 + 1/\beta).$ If you happen to work exclusively with special 
cases (called conjugate distributions) whose integrals you can look up in a 
textbook, then this is as far as you need to read.  But this strategy doesn't
work in general: if we have uncertainty about the values of $\alpha$ and 
$\beta$, or if the intensities come from some other distribution like the 
lognormal, or if we wanted to include additional prior information about 
$\lambda$, or if we wanted to integrate out $\lambda$ and another parameter
at the same time, we'd be stuck with something we couldn't compute.

In short, we'll have to use some kind of approximation pretty often.

# How to avoid integrating

## Numerical integration, part 1: Newton–Cotes

If you've taken a calculus class that covers integration, you might remember 
"the rectangle rule", "the trapezoid rule", and/or "Simpson's rule". These
methods all estimate the area under a curve by examining the height of the curve
(and possibly its slope and other derivatives) at a series of closely-spaced 
points. This gives an approximation of the curve's shape; calculating the area
under our approximation gives us an estimate of the area under the original 
curve.

This is rarely the best method to use in statistics (especially when integrating
over multiple dimensions at once), but it is one of the easiest ones to 
understand.

```{r}
delta = 0.5
lambda_sequence = seq(0, xmax, delta)
rect_lik = dpois(0, lambda_sequence) * dgamma(lambda_sequence, shape, rate)
rect_xmin = lambda_sequence - delta / 2
rect_xmax = lambda_sequence + delta / 2

plot(NULL, xlim = c(0, xmax), ylim = c(0, posterior_height))
rect(rect_xmin, 0, rect_xmax, rect_lik)
curve(dpois(0, x) * dgamma(x, shape, rate), add = TRUE)

sum(rect_lik) * delta
dnbinom(0, size = shape, prob = prob)
```


## Monte Carlo

This is of the most commonly-used approaches, especially in Bayesian statistics.
The underlying idea is straightforward: Instead using *all* possible values for 
$\lambda$, work with a representative sample.

Here, we know what the distribution of $\lambda$ is, so it's easy to generate
10 million representative samples and determine their Poisson likelihoods.
The answer we get this way agrees closely with the correct answer, generated
by the negative binomial distribution.

```{r}
N = 10000000
random_lambdas = rgamma(N, shape, rate)
mean(dpois(0, random_lambdas))
dnbinom(0, size = shape, prob = prob)
```

In practice, it's usally hard to sample from the parameter distribution,
and fancier versions of this approach are needed.

## Laplace approximation

```{r}
log_likelihood_log_scale = function(log_lambda) {
  log(likelihood(exp(log_lambda)))
}
mle_log = optimize(log_likelihood_log_scale, c(-10, 10), maximum = TRUE)$maximum
sigma_log = 1/sqrt(-numDeriv::hessian(log_likelihood_log_scale, mle_log))

log_lambda_sequence = seq(log(1/xmax), log(xmax), length = 1000)

laplace_amount = likelihood(exp(mle_log)) / dnorm(mle_log, mle_log, sigma_log)
laplace_amount2 = likelihood(exp(mle_log)) / dlnorm(exp(mle_log), mle_log, sigma_log)

curve(laplace_amount * dnorm(log(x), mle_log, sigma_log), add = FALSE, n = 1000, 
      from = .01, to = xmax, log = "x", ylab = "density",
      ylim = c(0, posterior_height))
curve(likelihood(x), add = TRUE, col = "blue")

# Natural scale
curve(laplace_amount2 * dlnorm(x, mle_log, sigma_log), add = FALSE, n = 1000, 
      from = .01, to = xmax, ylab = "density",
      ylim = c(0, posterior_height))
curve(likelihood(x), add = TRUE, col = "blue")


laplace_amount2
```

* Wald approximation, often used for margin of error
* Used by `glm` function for confidence intervals


## Numerical integration, part 2: Gaussian quadrature & Gauss-Hermite quadrature

```{r}
library(fastGHQuad)

sigma = 1/sqrt(-numDeriv::hessian(
  function(x)log_likelihood_log_scale(exp(x)), 
  exp(mle_log)))

ghquad = function(i){
  aghQuad(
    function(x){
      sapply(
        x,
        function(lambda){
          if (lambda > 0) {
            likelihood(lambda)
          }else{
            0
          }
        }
      )
    }, 
    muHat = exp(mle_log), 
    sigmaHat = sigma_log, 
    rule = gaussHermiteData(i))
}

plot(sapply(1:50, ghquad))
abline(h = dnbinom(0, size = shape, prob = prob))
```

## Variational approximation

## Numerical integration, part 3: Bayesian Quadrature



